Examining spatial equity and accessibility to a public bicycle share program using a balanced floating catchment area approach

### 2021-03-24 Bike-share-spatial-equity/Bike-share-spatial-equity.tex
Manuscript draft completed and sent to co-authors for review

Public bicycle share programs (PBSP) have been implemented in over 800
cities worldwide and a great deal has been learned about their typical
users ({\textbf{???}}). In many cities, males use bike share more than
females ({\textbf{???}}; {\textbf{???}}; {\textbf{???}}; {\textbf{???}})
as do younger age cohorts ({\textbf{???}}; {\textbf{???}};
{\textbf{???}}). However, one study found that bike share users in
Washington, DC were more likely to be female ({\textbf{???}}), which
suggests that the gender gap among cyclists who use bike share is less
disparate than the gap for personal bicycle use ({\textbf{???}}). There
is some evidence that bike share users are less likely to own a car
({\textbf{???}}; {\textbf{???}}). However, the relationship between
income or education and bike share use is less clear. Stations in
disadvantaged communities in Chicago generated most of the average
annual trips ({\textbf{???}}) and individuals from minority or lower
socioeconomic status neighbourhoods in Minneapolis-St.~Paul used the
city's PBSP more ({\textbf{???}}). Being university educated was also a
significant correlate of bike share use in Montreal, Canada
({\textbf{???}}). On the other hand, members who reside in
minority-concentrated and lower socioeconomic status neighbourhoods use
the Minneapolis-St Paul bike share more frequently ({\textbf{???}}).
Financial savings have been found to motivate those on a low income to
use bike share ({\textbf{???}}). Many studies have found that proximity
to a bike share station, either living or working near one, is an
important determinant of use or having a membership ({\textbf{???}};
{\textbf{???}}). This makes sense given that individuals are more likely
to use services or programs that they can easily access. Finally, the
potential of bike share programs to increase cycling levels has also
been explored recently ({\textbf{???}}; {\textbf{???}}), but more
longitudinal evidence is needed to determine their impact on encouraging
more cycling and offering more opportunities for physical activity. On
the whole, these findings highlight the need for bike share systems to
be more highly accessible for diverse populations in order to increase
use beyond the ``typical'' users, which has been the focus of recent
research ({\textbf{???}}; {\textbf{???}}; {\textbf{???}}), and to offer
more people the option of using sustainable and active transportation.

Several studies have recently explored the equity of PBSP in North
American cities by primarily examining who has access to bike share
(e.g., differences by demographics or socioeconomic status) and where
stations are located. Equity can be achieved in two different ways:
equal spatial distribution across a region (e.g., horizontal equity) or
greater access for vulnerable or disadvantaged populations (e.g.,
vertical equity) ({\textbf{???}}). Both are of interest to researchers
and transport planners since they are often linked in that advantage, or
conversely disadvantage, has spatial patterns. Using a negative binomial
regression model, Qian and Jaller ({\textbf{???}}) estimated ridership
in Chicago's PBSP among disadvantaged communities and found some
disparities. While annual members in disadvantaged communities have a
significantly lower share of trips compared to other areas in the city,
they make longer trips. This suggests that they may be using PBSP for
work commuting, which points to the importance of ensuring equitable
access ({\textbf{???}}). Similar results were found in Philadelphia,
where lower income areas generate fewer trips and efforts to increase
equity within the program have not been as successful as intended
({\textbf{???}}). In the case of Seattle, all neighbourhoods had some
level of access to dockless bikes but those with higher incomes and more
residents of higher education had more bikes ({\textbf{???}}). Babagoli
et al.~({\textbf{???}}) found that neighbourhoods in New York City with
higher affluence had the greater proportion of Citi Bike stations.
Overall, these findings suggest that horizontal equity can be achieved
while vertical equity is harder to attain. Exploring transport equity by
investigating where bicycle share stations are located, often using
neighbourhoods or census tracts as the spatial unit of analysis, can
ignore or miss the benefits that may be derived from adjacent zones.
Meaning that, stations may be lacking in certain neighbourhoods but
there may be stations accessible within a reasonable walking cutoff
time. This is where accessibility becomes an important consideration.

Accessibility has been applied in both a positive and normative way to
inform transportation planning ({\textbf{???}}), but its utility to this
field has evolved since its conceptualization and become linked with
recent interest in prioritizing local proximity and modes that are
suitable for local trips like walking and cycling ({\textbf{???}}). As a
measure of the ease of reaching potential destinations spread spatially
in a given area, accessibility is relevant to PBSPs because it can
identify current inequities in usage, as well as guide interventions
that increase access for groups that are under-serviced or address gaps
in transportation options. It also addresses some of the challenges of
other performance measures such as level of service within a
transportation network by measuring person-based indicators and
exploring differences in use between population groups ({\textbf{???}}).

In the context of PBSPs, accessibility refers to the distance an
individual must travel to reach a bicycle share station ({\textbf{???}};
{\textbf{???}}). Since the time or distance that members of the
population need to reach a bicycle share station decreases the potential
of accessing the program and ultimately use of the program, the location
and size (e.g., maximum number of bicycles available) of stations
matters. Indeed, distance to bicycle share stations is associated with
use ({\textbf{???}}; {\textbf{???}}) and can be a barrier to using PBSPs
({\textbf{???}}). Kabra et al.~({\textbf{???}}) found that the majority
of bike share usage in Paris comes from areas within 300m of stations.
Furthermore, unlike other public amenities with greater ability to adapt
to crowding, like health care services, the utility of stations is
limited by the number of bicycles that they can hold, which makes
crowding effects critical. The program may not necessarily be improved
if stations are easy to reach but offer only a small number of bicycles.
Likewise, more people may not opt to use the program if the supply of
bicycles available at the nearest station is insufficient to meet
demand. The location and size of stations is important to increase the
utility of this public transportation option for more people, thus
achieving vertical equity. Accessibility analyses for PBSPs constitute a
positive and evaluation-based approach that also has the potential to
inform equity efforts. For instance, Wang and Lindsey ({\textbf{???}})
investigated whether new or relocated bicycle share stations increased
accessibility and use, which offered important insights to improve the
performance of the program.

Several approaches have been commonly used for measuring accessibility:
cumulative opportunities, gravity, and utility-based ({\textbf{???}}).
Paez et al ({\textbf{???}}) provide a recent overview of various
formulations and applications of accessibility in transportation
research. Briefly, the gravity-based approach involves weighting
opportunities, for example the quantity of bicycle share stations, as
measured by a function of time, using a negative exponential form
({\textbf{???}}). It takes both demand and congestion into account
({\textbf{???}}). Floating catchment area (FCA) methods are a type of
gravity measure that have been used in healthcare accessibility research
but that are applicable to transportation case studies. This approach
involves producing flexible catchment areas for populations, recognizing
that people may be willing to travel to access particular services
({\textbf{???}}). Thus, this model does a good job of considering
potential crowding of services using a binary distance function
({\textbf{???}}). A recent improvement to this approach was achieved
through a simple and intuitive adjustment to the allocation of supply
and demand, which addressed the effects of demand and service inflation
from the conventional approach (see {\textbf{???}}).

Researchers have also taken different approaches when it comes to
aggregation of data, either by using the individual or household as the
smallest unit of analysis or larger spatial zones. Previous research on
bike share equity has typically used a macro-level approach with
aggregate data from entire neighbourhoods or census tracts
({\textbf{???}}; {\textbf{???}}; {\textbf{???}}; {\textbf{???}}),
although there are recent exceptions ({\textbf{???}}; {\textbf{???}}).
This is also true for studies examining correlates of bike share demand
({\textbf{???}}). Handy and Niemeier ({\textbf{???}}) note that using
disaggregated data in accessibility analyses provides a more accurate
estimate for individuals, which is useful for addressing vertical
inequities in PBSP usage.

@@

This paper uses the city of Hamilton, located in Ontario, Canada, as a
case study. The city launched a public bicycle share program, Hamilton
Bike Share, in March 2015 with 115 stations and 750 bicycles
({\textbf{???}}). Before June 2020, the program was known as SoBi
Hamilton. Stations are spaced between 300 and 600 metres apart
({\textbf{???}}). The core service area spans 35 km\^{}2 of the city and
roughly 138,000 can reach a bike share hub within 30 minutes of walking
{[}see Figure 1{]}. This represents roughly one fifth of the population
in Hamilton, according to the 2016 Canadian Census. The program was
enthusiastically welcomed in the city - within three weeks of launching,
10,000 trips had been made ({\textbf{???}}). In 2017, Hamilton Bike
Share Inc., the non-profit organization that operates the program,
initiated an equity program, Everyone Rides Initiative (ERI), to remove
barriers that may prevent individuals from accessing bike share in
Hamilton. An additional 75 bicycles and 15 stations were added to the
program, which expanded it to more disadvantaged areas in the core
service area {[}see Figure 2{]}. The program also offers subsidized
memberships to individuals who identify as low income. A comparable
program can be found in Philadelphia (see {\textbf{???}}). As of June
2020, the bike share program has 900 bikes and 130 stations, and over
26,000 active memberships ({\textbf{???}}).

Hamilton Bike Share has conducted one membership survey to date in 2018
({\textbf{???}}), and the findings from 420 members are broadly in line
with the trends that were discussed above (see {\textbf{???}} for a
recent review of the literature). The majority of respondents live
within the core service area and the gender split is typical: 57\% of
respondents are male and 41\% are female. The majority of respondents,
both male and female, are between 25 and 34 years of age, but the
percentage of male respondents is higher in the subsequent age groups.
Respondents use bike share for commuting (40\% of trips) or errands and
meetings (24\% of trips), and nearly 50\% of trips have an average
length of 11 to 20 minutes. As a result of having a bike share
membership, 49\% of respondents report that they use their private
vehicle less often or much less often but 48\% report that their private
vehicle use has remained about the same. This suggests that SoBi
Hamilton has been useful for certain kinds of trips but not all, meaning
that some trips continue to require a private vehicle.

Further research has provided insights about the behaviour and route
choice of bike share users in Hamilton. The routes most frequently
travelled are longer than the shortest distance route from origin to
destination ({\textbf{???}}; {\textbf{???}}). Ridership is influenced by
weather conditions, temporal factors such as university terms,
employment, and proximity to important destinations like the local
university, but is not influenced by population and most transportation
infrastructure variables ({\textbf{???}}). Bike share users prefer to
deter from the shortest distance route to travel on streets with cycling
facilities or lower volumes of traffic ({\textbf{???}}) and avoid steep
slopes and busy roads ({\textbf{???}}). It is important to note that
these studies used daily ridership or GPS data from trips taken before
the equity stations were implemented, so these findings are informative
with respect to the conventional stations.

Our analysis builds upon a previous and recent study ({\textbf{???}}),
which found that disadvantaged areas in Hamilton are better served by
the city's PBSP compared to other Canadian cities {[}i.e., Toronto,
Vancouver, Montreal, and Ottawa-Gatineau{]} with PBSPs where advantaged
areas have greater access. Hosford and Winters ({\textbf{???}})
acknowledge that ``Hamilton stands out in that the lower income
neighborhoods are located near the city center and wealthier
neighborhoods are in the surrounding suburban areas''. Therefore, the
core service area for the PBSP in Hamilton is by default in more
disadvantaged areas compared to other Canadian cities, but there is also
a great deal of variation in income in the city center as well because
of the local university and increasing gentrification in the downtown
area. In their analysis, Hosford and Winters ({\textbf{???}}) did not
differentiate between the conventional stations and those added to
increase equity, which we refer to as equity stations in this paper.
Therefore, this paper continues their work and makes an important
contribution to the growing body of literature that examines equity
issues in PBSPs.

@@

\caption{The core service area of Hamilton Bike Share is outlined in blue. Dissemination areas in the Hamilton Census Metropolitan Area (CMA) are outlined in grey.}\label{fig:hamilton-and-sobi-service-area}
\end{figure}

@@

\caption{The spatial distribution of bike share stations in Hamilton, Ontario. The service area of Hamilton Bike Share is outlined in blue and the city's downtown core is outlined in light blue. Dissemination areas within the Hamilton CMA that touch the bounding box of the core service area are outlined in grey.}\label{fig:sobi-stations-in-hamilton}
\end{figure}

@@

Floating catchment area (FCA) methods are an approach commonly used in
the healthcare accessibility literature. This approach is more
appropriate and informative than calculating the provider-to-population
ratio (PPR) which simply divides the level of supply of a service (e.g.,
the number of bicycle racks at a station) by the population who have
access to the service ({\textbf{???}}). In particular, the Two-Step
Floating Catchment Area method produces flexible catchment areas for
populations accessing a service rather than using rigid boundaries like
PPR, which provides more useful information because it takes into
account individual behaviour and doesn't assume that people stay within
the pre-defined boundaries ({\textbf{???}}). This is an important
property that supports our rationale for applying this method to measure
accessibility to Hamilton Bike Share. The City of Hamilton has
positioned stations between 300 and 600 metres apart, which would lead
to boundaries around these buffers, but it would be reasonable to assume
that people are willing to walk beyond this threshold to access other
stations if the ones nearest them have no supply of bicycles.

More recently, the \emph{balanced} floating catchment area approach was
developed to address issues with demand and supply inflation that are a
result of the overlapping catchment areas produced by the Two-Step
Floating Catchment Area ({\textbf{???}}). This overlap inflates the
demand and generates inaccurate or misleading accessibility estimates
({\textbf{???}}). By adjusting the impedance weights, both supply and
demand are proportionally allocated which eliminates the double-counting
of population that leads to demand and supply inflation
({\textbf{???}}). Other benefits of this adjusted method include
consideration of competition which can occur when catchment areas
overlap, as well as the preservation of system-wide population and level
of service. Balanced floating catchment area methods have been used to
explore accessibility to health care providers ({\textbf{???}}) and
COVID-19 health care services ({\textbf{???}}), but have not yet been
used in the cycling literature to explore issues of accessibility.

In their review of accessibility measures, Geurs and van Wee
({\textbf{???}}) highlight the need for greater inclusion of individual
spatio-temporal constraints but acknowledge the challenges of acquiring
and analyzing person-based data. This comes after Kwan's
({\textbf{???}}) work to show that space-time measures are more capable
of capturing interpersonal differences, especially the effect of
space-time constraints on individual behaviour, and are more helpful for
unraveling gender/ethnic differences. Applying the balanced floating
catchment area approach to examine accessibility to Hamilton Bike Share
constitutes a location-based measure with an individual component by
stratifying according to median household income ({\textbf{???}}), but
conducting a further sensitivity analysis by adjusting distance
thresholds would introduce additional spatio-temporal constraints to
evaluate equitable accessibility.

@@

We first plotted the population in all dissemination areas within SoBi
Hamilton's core service area. Since pycnophylactic interpolation occurs
at such a micro scale, we had to ensure that population numbers were not
interpolated or allocated to areas where people do not live in Hamilton
(i.e., where schools or parks are located). To do so, we retrieved
shapefiles for various geographic features (see Table 1) from Open
Hamilton. Figure 3 shows where these spaces are located within SoBi
Hamilton's core service area. Next, we extracted these features from the
PBSP core service area and used pycnophylactic interpolation to
disaggregate and reallocate population from dissemination areas to
smaller polygons. These micro-level polygons are 50 by 50 metres.

@@

\caption{\label{tab:data-features}\label{tab:landscape-features}Landscape features}
\centering
\resizebox{\linewidth}{!}{
\begin{tabular}[t]{>{}l|>{\raggedright\arraybackslash}p{30em}}
\toprule
Feature & Description\\
\midrule
\cellcolor{gray!6}{\textbf{Hamilton Bike Share Stations}} & \cellcolor{gray!6}{The location of stations and the number of racks available at each station.}\\
\textbf{Golf Courses} & The location of City and privately owned golf courses.\\
\cellcolor{gray!6}{\textbf{Parks}} & \cellcolor{gray!6}{The location of parks and other green spaces.}\\
\textbf{Designated Large Employment Areas} & The location of large business parks and industrial lands.\\
\cellcolor{gray!6}{\textbf{Municipal Parking Lots}} & \cellcolor{gray!6}{The location of municipal car parks.}\\
\addlinespace
\textbf{Cemeteries} & The location of cemeteries.\\
\cellcolor{gray!6}{\textbf{Environmentally Sensitive Areas}} & \cellcolor{gray!6}{The location of either land or water areas containing natural features or significant ecological functions.}\\
\textbf{Streets} & The street network in Hamilton, including road classification for highways.\\
\cellcolor{gray!6}{\textbf{Educational Institutions}} & \cellcolor{gray!6}{The location of all educational institutions and schools.}\\
\textbf{Places of Worship} & The location of buildings used for religious congregations.\\
\addlinespace
\cellcolor{gray!6}{\textbf{Municipal Service Centres}} & \cellcolor{gray!6}{The location of all municipal service centres, including City Hall.}\\
\textbf{Recreation and Community Centres} & The location of all recreation and community centres.\\
\cellcolor{gray!6}{\textbf{Arenas}} & \cellcolor{gray!6}{The location of all indoor arenas.}\\
\textbf{Emergency Stations} & The location of all Emergency Management Services (EMS) Ambulance stations.\\
\cellcolor{gray!6}{\textbf{Fire Stations}} & \cellcolor{gray!6}{The location of all fire stations.}\\
\addlinespace
\textbf{Police Stations} & The location of all police stations.\\
\cellcolor{gray!6}{\textbf{Railways}} & \cellcolor{gray!6}{The railway network in Hamilton.}\\
\textbf{Hospitals} & The location of all hospitals.\\
\bottomrule
\end{tabular}}
\end{table}

@@

{\centering \includegraphics[width=0.65\linewidth]{Bike-share-spatial-equity_files/figure-latex/figure-2-1}

\caption{Population in all dissemination areas in the SoBi Hamilton core service area and within 30 minutes of walking to a bike share station.}\label{fig:figure-2}
\end{figure}

{\centering \includegraphics[width=0.65\linewidth]{Bike-share-spatial-equity_files/figure-latex/figure-4-1}

\caption{Interpolated population cells in the SoBi Hamilton core service area (outlined in black) and within 30 minutes of walking to the core service area (outlined in blue). The downtown area is outlined in dark green. Geographic features have been extracted. Dissemination area polygons are outlined in grey.}\label{fig:figure-4}
\end{figure}

>>>>

Public bicycle share programs (PBSP) have been implemented in over 800
cities worldwide and a great deal has been learned about their typical
users (Fishman, 2016). In many cities, males use bike share more than
females (Brey et al., 2017; Nickkar et al., 2019; Reilly, S. M. Wang, et
al., 2020; Winters et al., 2019) as do younger age cohorts (Brey et al.,
2017; Buck et al., 2013; Fuller et al., 2011). However, one study found
that bike share users in Washington, DC were more likely to be female
(Buck et al., 2013), which suggests that the gender gap among cyclists
who use bike share is less disparate than the gap for personal bicycle
use (Fishman, 2016). There is some evidence that bike share users are
less likely to own a car (Buck et al., 2013; Reilly, Noyes, et al.,
2020). However, the relationship between income or education and bike
share use is less clear. Stations in disadvantaged communities in
Chicago generated most of the average annual trips (Qian and Jaller,
2020) and individuals from minority or lower socioeconomic status
neighbourhoods in Minneapolis-St.~Paul used the city's PBSP more (J.
Wang and Lindsey, 2019a). Being university educated was also a
significant correlate of bike share use in Montreal, Canada (Fuller et
al., 2011). On the other hand, members who reside in
minority-concentrated and lower socioeconomic status neighbourhoods use
the Minneapolis-St Paul bike share more frequently (J. Wang and Lindsey,
2019a). Financial savings have been found to motivate those on a low
income to use bike share (Fishman, 2016). Many studies have found that
proximity to a bike share station, either living or working near one, is
an important determinant of use or having a membership (Fishman, 2016;
Fuller et al., 2011). This makes sense given that individuals are more
likely to use services or programs that they can easily access. Finally,
the potential of bike share programs to increase cycling levels has also
been explored recently (Hosford et al., 2018, 2019), but more
longitudinal evidence is needed to determine their impact on encouraging
more cycling and offering more opportunities for physical activity. On
the whole, these findings highlight the need for bike share systems to
be more highly accessible for diverse populations in order to increase
use beyond the ``typical'' users, which has been the focus of recent
research (Auchincloss et al., 2020; Hull Grasso et al., 2020; MacArthur
et al., 2020), and to offer more people the option of using sustainable
and active transportation.

Several studies have recently explored the equity of PBSP in North
American cities by primarily examining who has access to bike share
(e.g., differences by demographics or socioeconomic status) and where
stations are located. Equity can be achieved in two different ways:
equal spatial distribution across a region (e.g., horizontal equity) or
greater access for vulnerable or disadvantaged populations (e.g.,
vertical equity) (Chen et al., 2019). Both are of interest to
researchers and transport planners since they are often linked in that
advantage, or conversely disadvantage, has spatial patterns. Using a
negative binomial regression model, Qian and Jaller (2020) estimated
ridership in Chicago's PBSP among disadvantaged communities and found
some disparities. While annual members in disadvantaged communities have
a significantly lower share of trips compared to other areas in the
city, they make longer trips. This suggests that they may be using PBSP
for work commuting, which points to the importance of ensuring equitable
access (Qian and Jaller, 2020). Similar results were found in
Philadelphia, where lower income areas generate fewer trips and efforts
to increase equity within the program have not been as successful as
intended (Caspi and Noland, 2019). In the case of Seattle, all
neighbourhoods had some level of access to dockless bikes but those with
higher incomes and more residents of higher education had more bikes
(Mooney et al., 2019). Babagoli et al.~(2019) found that neighbourhoods
in New York City with higher affluence had the greater proportion of
Citi Bike stations. Overall, these findings suggest that horizontal
equity can be achieved while vertical equity is harder to attain.
Exploring transport equity by investigating where bicycle share stations
are located, often using neighbourhoods or census tracts as the spatial
unit of analysis, can ignore or miss the benefits that may be derived
from adjacent zones. Meaning that, stations may be lacking in certain
neighbourhoods but there may be stations accessible within a reasonable
walking cutoff time. This is where accessibility becomes an important
consideration.

Accessibility has been applied in both a positive and normative way to
inform transportation planning (Páez et al., 2012), but its utility to
this field has evolved since its conceptualization and become linked
with recent interest in prioritizing local proximity and modes that are
suitable for local trips like walking and cycling (Levine, 2020). As a
measure of the ease of reaching potential destinations spread spatially
in a given area, accessibility is relevant to PBSPs because it can
identify current inequities in usage, as well as guide interventions
that increase access for groups that are under-serviced or address gaps
in transportation options. It also addresses some of the challenges of
other performance measures such as level of service within a
transportation network by measuring person-based indicators and
exploring differences in use between population groups (Páez et al.,
2012).

In the context of PBSPs, accessibility refers to the distance an
individual must travel to reach a bicycle share station (Kabra et al.,
2020; J. Wang and Lindsey, 2019b). Since the time or distance that
members of the population need to reach a bicycle share station
decreases the potential of accessing the program and ultimately use of
the program, the location and size (e.g., maximum number of bicycles
available) of stations matters. Indeed, distance to bicycle share
stations is associated with use (Fuller et al., 2011; J. Wang and
Lindsey, 2019b) and can be a barrier to using PBSPs (Fishman et al.,
2014). Kabra et al.~(2020) found that the majority of bike share usage
in Paris comes from areas within 300m of stations. Furthermore, unlike
other public amenities with greater ability to adapt to crowding, like
health care services, the utility of stations is limited by the number
of bicycles that they can hold, which makes crowding effects critical.
The program may not necessarily be improved if stations are easy to
reach but offer only a small number of bicycles. Likewise, more people
may not opt to use the program if the supply of bicycles available at
the nearest station is insufficient to meet demand. The location and
size of stations is important to increase the utility of this public
transportation option for more people, thus achieving vertical equity.
Accessibility analyses for PBSPs constitute a positive and
evaluation-based approach that also has the potential to inform equity
efforts. For instance, Wang and Lindsey (2019b) investigated whether new
or relocated bicycle share stations increased accessibility and use,
which offered important insights to improve the performance of the
program.

Several approaches have been commonly used for measuring accessibility:
cumulative opportunities, gravity, and utility-based (Handy and
Niemeier, 1997). Paez et al (2012) provide a recent overview of various
formulations and applications of accessibility in transportation
research. Briefly, the gravity-based approach involves weighting
opportunities, for example the quantity of bicycle share stations, as
measured by a function of time, using a negative exponential form (Handy
and Niemeier, 1997). It takes both demand and congestion into account
(Páez et al., 2012). Floating catchment area (FCA) methods are a type of
gravity measure that have been used in healthcare accessibility research
but that are applicable to transportation case studies. This approach
involves producing flexible catchment areas for populations, recognizing
that people may be willing to travel to access particular services (Páez
et al., 2012). Thus, this model does a good job of considering potential
crowding of services using a binary distance function (Páez et al.,
2012). A recent improvement to this approach was achieved through a
simple and intuitive adjustment to the allocation of supply and demand,
which addressed the effects of demand and service inflation from the
conventional approach (see Paez et al., 2019).

Researchers have also taken different approaches when it comes to
aggregation of data, either by using the individual or household as the
smallest unit of analysis or larger spatial zones. Previous research on
bike share equity has typically used a macro-level approach with
aggregate data from entire neighbourhoods or census tracts (Babagoli et
al., 2019; Mooney et al., 2019; Qian and Jaller, 2020; J. Wang and
Lindsey, 2019a), although there are recent exceptions (Chen et al.,
2019; Chen and Li, 2021). This is also true for studies examining
correlates of bike share demand (J. Wang and Lindsey, 2019b). Handy and
Niemeier (1997) note that using disaggregated data in accessibility
analyses provides a more accurate estimate for individuals, which is
useful for addressing vertical inequities in PBSP usage.

@@

This paper uses the city of Hamilton, located in Ontario, Canada, as a
case study. The city launched a public bicycle share program, Hamilton
Bike Share, in March 2015 with 115 stations and 750 bicycles (Hamilton,
2015). Before June 2020, the program was known as SoBi Hamilton.
Stations are spaced between 300 and 600 metres apart (Scott et al.,
2021). The core service area spans 35 km\^{}2 of the city and roughly
138,000 can reach a bike share hub within 30 minutes of walking {[}see
Figure \ref{fig:hamilton-and-sobi-service-area}{]}. This represents
roughly one fifth of the population in Hamilton, according to the 2016
Canadian Census. The program was enthusiastically welcomed in the city -
within three weeks of launching, 10,000 trips had been made (Hamilton,
2015). In 2017, Hamilton Bike Share Inc., the non-profit organization
that operates the program, initiated an equity program, Everyone Rides
Initiative (ERI), to remove barriers that may prevent individuals from
accessing bike share in Hamilton. An additional 75 bicycles and 15
stations were added to the program, which expanded it to more
disadvantaged areas in the core service area {[}see Figure 2{]}. The
program also offers subsidized memberships to individuals who identify
as low income. A comparable program can be found in Philadelphia (see
Caspi and Noland, 2019). As of June 2020, the bike share program has 900
bikes and 130 stations {[}see Figure
\ref{fig:sobi-stations-in-hamilton}{]}, and over 26,000 active
memberships (Hamilton, 2015).

Hamilton Bike Share has conducted one membership survey to date in 2018
(Civicplan, 2017), and the findings from 420 members are broadly in line
with the trends that were discussed above (see Fishman, 2016 for a
recent review of the literature). The majority of respondents live
within the core service area and the gender split is typical: 57\% of
respondents are male and 41\% are female. The majority of respondents,
both male and female, are between 25 and 34 years of age, but the
percentage of male respondents is higher in the subsequent age groups.
Respondents use bike share for commuting (40\% of trips) or errands and
meetings (24\% of trips), and nearly 50\% of trips have an average
length of 11 to 20 minutes. As a result of having a bike share
membership, 49\% of respondents report that they use their private
vehicle less often or much less often but 48\% report that their private
vehicle use has remained about the same. This suggests that SoBi
Hamilton has been useful for certain kinds of trips but not all, meaning
that some trips continue to require a private vehicle.

Further research has provided insights about the behaviour and route
choice of bike share users in Hamilton. The routes most frequently
travelled are longer than the shortest distance route from origin to
destination (Lu et al., 2018; Scott et al., 2021). Ridership is
influenced by weather conditions, temporal factors such as university
terms, employment, and proximity to important destinations like the
local university, but is not influenced by population and most
transportation infrastructure variables (Scott and Ciuro, 2019). Bike
share users prefer to deter from the shortest distance route to travel
on streets with cycling facilities or lower volumes of traffic (Lu et
al., 2018) and avoid steep slopes and busy roads (Scott et al., 2021).
It is important to note that these studies used daily ridership or GPS
data from trips taken before the equity stations were implemented, so
these findings are informative with respect to the conventional
stations.

Our analysis builds upon a previous and recent study (Hosford and
Winters, 2018), which found that disadvantaged areas in Hamilton are
better served by the city's PBSP compared to other Canadian cities
{[}i.e., Toronto, Vancouver, Montreal, and Ottawa-Gatineau{]} with PBSPs
where advantaged areas have greater access. Hosford and Winters (2018)
acknowledge that ``Hamilton stands out in that the lower income
neighborhoods are located near the city center and wealthier
neighborhoods are in the surrounding suburban areas''. Therefore, the
core service area for the PBSP in Hamilton is by default in more
disadvantaged areas compared to other Canadian cities, but there is also
a great deal of variation in income in the city center as well because
of the local university and increasing gentrification in the downtown
area. In their analysis, Hosford and Winters (2018) did not
differentiate between the conventional stations and those added to
increase equity, which we refer to as equity stations in this paper.
Therefore, this paper continues their work and makes an important
contribution to the growing body of literature that examines equity
issues in PBSPs.

@@

\caption{The core service area of Hamilton Bike Share is outlined in black. Hamilton Census Metropolitan Area (CMA) is shown in grey.}\label{fig:hamilton-and-sobi-service-area}
\end{figure}

@@

\caption{The spatial distribution of bike share stations in Hamilton, Ontario. The service area of Hamilton Bike Share is outlined in black and the city's downtown core is outlined in dark green. Hamilton CMA is shown in grey.}\label{fig:sobi-stations-in-hamilton}
\end{figure}

@@

Floating catchment area (FCA) methods are an approach commonly used in
the healthcare accessibility literature. This approach is more
appropriate and informative than calculating the provider-to-population
ratio (PPR) which simply divides the level of supply of a service (e.g.,
the number of bicycle racks at a station) by the population who have
access to the service (Paez et al., 2019). In particular, the Two-Step
Floating Catchment Area method produces flexible catchment areas for
populations accessing a service rather than using rigid boundaries like
PPR, which provides more useful information because it takes into
account individual behaviour and doesn't assume that people stay within
the pre-defined boundaries (Paez et al., 2019). This is an important
property that supports our rationale for applying this method to measure
accessibility to Hamilton Bike Share. The City of Hamilton has
positioned stations between 300 and 600 metres apart, which would lead
to boundaries around these buffers, but it would be reasonable to assume
that people are willing to walk beyond this threshold to access other
stations if the ones nearest them have no supply of bicycles.

More recently, the \emph{balanced} floating catchment area approach was
developed to address issues with demand and supply inflation that are a
result of the overlapping catchment areas produced by the Two-Step
Floating Catchment Area (Paez et al., 2019). This overlap inflates the
demand and generates inaccurate or misleading accessibility estimates
(Paez et al., 2019). By adjusting the impedance weights, both supply and
demand are proportionally allocated which eliminates the double-counting
of population that leads to demand and supply inflation (Paez et al.,
2019). Other benefits of this adjusted method include consideration of
competition which can occur when catchment areas overlap, as well as the
preservation of system-wide population and level of service. Balanced
floating catchment area methods have been used to explore accessibility
to health care providers (Paez et al., 2019) and COVID-19 health care
services (Pereira et al., 2021), but have not yet been used in the
cycling literature to explore issues of accessibility.

In their review of accessibility measures, Geurs and van Wee (2004)
highlight the need for greater inclusion of individual spatio-temporal
constraints but acknowledge the challenges of acquiring and analyzing
person-based data. This comes after Kwan's (1998) work to show that
space-time measures are more capable of capturing interpersonal
differences, especially the effect of space-time constraints on
individual behaviour, and are more helpful for unraveling gender/ethnic
differences. Applying the balanced floating catchment area approach to
examine accessibility to Hamilton Bike Share constitutes a
location-based measure with an individual component by stratifying
according to median household income (Geurs and van Wee, 2004), but
conducting a further sensitivity analysis by adjusting distance
thresholds would introduce additional spatio-temporal constraints to
evaluate equitable accessibility.

@@

Following the method first presented by Tobler (1979), we used
pycnophylactic interpolation to disaggregate population from each
dissemination area to smaller polygons that are 50 by 50 metres. This
method involves smoothing out the population from each dissemination
area while preserving total volume {[}see Figure
\ref{fig:da-population}{]}. Since pycnophylactic interpolation occurs at
such a micro scale, we had to ensure that population numbers were not
reallocated or smoothed out to areas where people do not live in
Hamilton (i.e., where schools or parks are located). To do so, we
retrieved shapefiles for various geographic features {[}see Table
\ref{tab:data-features}{]} from Open Hamilton. Next, we extracted these
features from the PBSP core service area and used pycnophylactic
interpolation to disaggregate and reallocate population {[}see Figure
\ref{fig:interpolated-population}{]}.

@@

\caption{\label{tab:data-features}\label{tab:landscape-features}Landscape features extracted from the Hamilton Bike Share core service area before population is interpolated.}
\centering
\resizebox{\linewidth}{!}{
\begin{tabular}[t]{>{}l|>{\raggedright\arraybackslash}p{30em}}
\toprule
Feature & Description\\
\midrule
\cellcolor{gray!6}{\textbf{Hamilton Bike Share Stations}} & \cellcolor{gray!6}{The location of stations and the number of racks available at each station.}\\
\textbf{Golf Courses} & The location of City and privately owned golf courses.\\
\cellcolor{gray!6}{\textbf{Parks}} & \cellcolor{gray!6}{The location of parks and other green spaces.}\\
\textbf{Designated Large Employment Areas} & The location of large business parks and industrial lands.\\
\cellcolor{gray!6}{\textbf{Municipal Parking Lots}} & \cellcolor{gray!6}{The location of municipal car parks.}\\
\addlinespace
\textbf{Cemeteries} & The location of cemeteries.\\
\cellcolor{gray!6}{\textbf{Environmentally Sensitive Areas}} & \cellcolor{gray!6}{The location of either land or water areas containing natural features or significant ecological functions.}\\
\textbf{Streets} & The street network in Hamilton, including road classification for highways.\\
\cellcolor{gray!6}{\textbf{Educational Institutions}} & \cellcolor{gray!6}{The location of all educational institutions and schools.}\\
\textbf{Places of Worship} & The location of buildings used for religious congregations.\\
\addlinespace
\cellcolor{gray!6}{\textbf{Municipal Service Centres}} & \cellcolor{gray!6}{The location of all municipal service centres, including City Hall.}\\
\textbf{Recreation and Community Centres} & The location of all recreation and community centres.\\
\cellcolor{gray!6}{\textbf{Arenas}} & \cellcolor{gray!6}{The location of all indoor arenas.}\\
\textbf{Emergency Stations} & The location of all Emergency Management Services (EMS) Ambulance stations.\\
\cellcolor{gray!6}{\textbf{Fire Stations}} & \cellcolor{gray!6}{The location of all fire stations.}\\
\addlinespace
\textbf{Police Stations} & The location of all police stations.\\
\cellcolor{gray!6}{\textbf{Railways}} & \cellcolor{gray!6}{The railway network in Hamilton.}\\
\textbf{Hospitals} & The location of all hospitals.\\
\bottomrule
\end{tabular}}
\end{table}

{\centering \includegraphics[width=0.65\linewidth]{Bike-share-spatial-equity_files/figure-latex/da-population-1}

\caption{Population in all dissemination areas (outlined in white) in the Hamilton Bike Share core service area (outlined in black) and within 30 minutes of walking to the service area.}\label{fig:da-population}
\end{figure}

{\centering \includegraphics[width=0.65\linewidth]{Bike-share-spatial-equity_files/figure-latex/interpolated-population-1}

\caption{Interpolated population in the Hamilton Bike Share core service area (outlined in black) and within 30 minutes of walking to the core service area (outlined in blue). The downtown area is outlined in dark green. Geographic features have been extracted. Dissemination area polygons are outlined in grey.}\label{fig:interpolated-population}
\end{figure}

@@
